Network Visualizations in \MakeLowercase{\texttt{ggplot2}}

### 2015-10-07 inst/paper.Rnw
added Francois
paper structure

\title{A Geometry for Network Visualization in \MakeLowercase{\texttt{ggplot2}}}
\author{Sam Tyner and Heike Hofmann}
\date{} % delete this line to display the current date
%%% BEGIN DOCUMENT
\begin{document}

@@

\par The three necessary elements of any network visualization are the vertices, the edges, and the layout.   \\
%But once those three items are visualized, we usually find our visualization to be lacking. \hh{pull it over to a positive statement}  We may want to color the vertices (or edges) by some sort of grouping variable, or we may want to make vertices of degree ten twice the size of vertices of degree five.

\hh{XXXX not sure where this next paragraph should go - but we need a more detailed discussion of ggnet anyways.}
\par The \texttt{GGally} package by \citet{ggally} contains a very useful function written by Fran\c{c}ois Briatte and Moritz Marbach called \texttt{ggnet} that does allow for fairly straightforward customization of these three necessary graph attributes.
 The \texttt{ggnet} function, however, requires the graph input to be a \texttt{network} object according to the \texttt{network} package.  Our work builds off of the \texttt{ggnet} function and presents an intuitive \texttt{geom} for network visualization within the \texttt{ggplot2} framework, without the need for objects other than simple data frames.
 \hh{XXXX}

\section{Two approaches of network visualizations}

\hh{this section needs a better name, but we need it, because here is where your work starts. it should not be part of the introduction.}
  \subsection{Data Structure}
In order for this geometry to work using data frames, there need to be two separate data frames given to the \texttt{geom\_net} function: one for the edge information and one for the vertex information. The vertex data frame should contain all the relevant vertex information. The only necessary variable is one called \texttt{label} which contains the vertex labels as they are encoded in the edges data set. This will be a formal requirement in the final version of the \texttt{geom\_net} function.  Each row represents one vertex.  Other values of interest, such as grouping variables or degree of each vertex should be stored as columns in this dataframe, with an observation for each vertex.
  \par The edge data frame should contain all the relevant edge information. The only necessary variables are the ``from vertex" and ``to vertex" for each edge in the network. The From and To vertices should match the names of the vertices in the vertex information data frame. The formal names for these columns are \texttt{from\_id} and \texttt{to\_id}, respectively.  Other variables may also be included for each edge, such as the edge weight or grouping variable.  As before, the variables of interest are columns in the data frame and the rows are each edge in the network.

  \subsection{Vertex Aesthetics}

@@

\par Evidently, we created all of these aesthetics to fit in exactly to the \texttt{ggplot2} grammar of graphics. We also created the edge aesthetics in a similar fashion.
  \subsection{Edge Aesthetics}
  Again, our edge aesthetics mimic the familiar \texttt{ggplot2} aesthetics, this time for segments.    The \texttt{ecolor} aesthetic changes the color of all of the edges according to the identity function or can change the color according to HEX colors assigned to levels of a grouping variable.  In the final version, the \texttt{ecolor} aesthetic will automatically color according to levels of a grouping variable or according to a gradient scale associated with a numeric variable.
  The \texttt{elinetype} aesthetic can also be used with a grouping variable to change the line type of each edge in the graph from solid to one of the different types of lines in \texttt{ggplot2}. Currently, the column associated with this aesthetic must be integer valued and take on values from zero to six. In the final version, it will change the edge linetypes according to levels of a grouping variable.  The \texttt{esize} aesthetic can also be set to the weight or probability of each edge, or it can be changed for all edges with the identity function. The associated column can take on any numerical values, and currently maps the values directly to points, just like the \texttt{vsize} aesthetic. In the final version, this aesthetic will work exactly like the sizing in \texttt{ggplot2}, by scaling sizes to the minimum and maximum values. This is one of two ways to visualize the number of edges between two vertices.  The other way is through the \texttt{ealpha} aesthetic. The \texttt{ealpha} aesthetic can change the edge transparency to a set value on the interval $[0,1]$.  If there are multiple edge connections given between two vertices in the edge data frame, the \texttt{ealpha} aesthetic can be changed to plot opaque lines between vertices with many mutual connections and varying degrees of transparent lines between vertices with only a few or a single connection. This property is also especially useful for large networks with a lot of connections or with vertices of relatively high degree. Finally, we will add an \texttt{elabel} aesthetic to print the name or number assigned to each edge on or next to it.  This aesthetic will be very useful for visualizing random graphs or Markov processes where each edge probability is of interest.

\subsection{Other Arguments}
  Similar to the other geometries and functions in \texttt{ggplot2}, our geometry takes several arguments outside of the aesthetic mapping parameters.  One such argument is the \texttt{vlabel} argument. The \texttt{vlabel} aesthetic is a logical value that, if set to \texttt{TRUE} prints the name or number associated with each vertex in the \texttt{label} column in order to identify it on the plot.  In the final version of this geometry, we will add capabilities to adjust the size and positioning of the labels. For the moment, the color is set to the vertex color, the size is set to twice the vertex size, and all other possible parameters are set to their defaults in \texttt{geom\_text()}. The \texttt{vlabel} argument is most useful for smaller network objects where all vertex names can be printed on their corresponding vertices and still be read clearly.  Next, the \texttt{directed} argument is a logical value that identifies whether or not the network is directed.  When this value is true, arrows are created on the end of the line segment that corresponds to the \texttt{to\_id} value for each edge.  The arrows match the edges in coloring, linetype, and size, and the default length of the arrow is set using the \texttt{unit} function in the \texttt{grid} package to 0.015 of the normalised parent coordinates of the plot.
  Finally, the \texttt{layout} argument takes a character value corresponding to the possible layouts in the \texttt{sna} package that are created by the \texttt{gplot.layout.*()} family of functions.  The default layout is the Kamada-Kawai layout.  This is a force-directed layout for undirected networks \citep{kamadakawai}. There are, however, many other layouts possible.
\begin{table}
\centering
\begin{tabular}{l|p{10cm}}
Layout Name & Description \\
\hline \hline
\texttt{"adj"} & a version of \texttt{"mds"} which scales the raw adjacency matrix \\
\texttt{"circle"}* & places vertices uniformly in a circle \\
\texttt{"circrand"} & places vertices randomly in a circle \\
\texttt{"eigen"} & places vertices based on the eigenstructure of the adjacency matrix \\
\texttt{"fruchtermanreingold"} & uses a variant of Fruchterman and Reingold's force-directed placement algorithm \\
\texttt{"geodist"} & a version of \texttt{"mds"} which scales the matrix of geodesic distances \\
\texttt{"hall"}* &  places vertices based on the last two eigenvectors of the Laplacian of the input matrix \\
\texttt{"kamadakawai"} & generates a vertex layout using a version of the Kamada-Kawai force-directed placement algorithm \\
\texttt{"mds"}  & places vertices based on a metric multidimensional scaling of a specified distance matrix \\
\texttt{"random"} & places vertices randomly \\
\texttt{"segeo"} & a version of \texttt{"mds"} which scales the squared euclidean distances between row-wise geodesic distances \\
\texttt{"seham"} & a version of \texttt{"mds"} which scales the Hamming distance between rows/columns of the adjacency matrix \\
\texttt{"target"} & produces a ``target diagram" or ``bullseye" layout using a force-directed placement algorithm.
\end{tabular}
\caption{All possible layouts to pass to \texttt{geom\_net}.  All descriptions are from \citet{sna}.  Note that the layouts marked with an asterisk (*) are the only layouts that do not take layout parameters.}
\label{table:layouts}
\end{table}

>>>>

\title{Network Visualizations in \MakeLowercase{\texttt{ggplot2}}}
\author{Sam Tyner and Fran\c{c}ois Briatte and  Heike Hofmann}
\date{} % delete this line to display the current date
%%% BEGIN DOCUMENT
\begin{document}

@@

\section{Two approaches of network visualizations}
 The three necessary elements of any network visualization are the vertices, the edges, and the layout.
%But once those three items are visualized, we usually find our visualization to be lacking. \hh{pull it over to a positive statement}  We may want to color the vertices (or edges) by some sort of grouping variable, or we may want to make vertices of degree ten twice the size of vertices of degree five.

\subsection{ggnet}
\hh{XXX needs massive changes :)}
\par The \texttt{GGally} package by \citet{ggally} contains a very useful function written by Fran\c{c}ois Briatte and Moritz Marbach called \texttt{ggnet} that does allow for fairly straightforward customization of these three necessary graph attributes.
 The \texttt{ggnet} function, however, requires the graph input to be a \texttt{network} object according to the \texttt{network} package.  Our work builds off of the \texttt{ggnet} function and presents an intuitive \texttt{geom} for network visualization within the \texttt{ggplot2} framework, without the need for objects other than simple data frames.
 \hh{XXXX}

\subsection{geom\_net}
  \subsubsection{Data Structure}
In order for this geometry to work using data frames, there need to be two separate data frames given to the \texttt{geom\_net} function: one for the edge information and one for the vertex information. The vertex data frame should contain all the relevant vertex information. The only necessary variable is one called \texttt{label} which contains the vertex labels as they are encoded in the edges data set. This will be a formal requirement in the final version of the \texttt{geom\_net} function.  Each row represents one vertex.  Other values of interest, such as grouping variables or degree of each vertex should be stored as columns in this dataframe, with an observation for each vertex.
  \par The edge data frame should contain all the relevant edge information. The only necessary variables are the ``from vertex" and ``to vertex" for each edge in the network. The From and To vertices should match the names of the vertices in the vertex information data frame. The formal names for these columns are \texttt{from\_id} and \texttt{to\_id}, respectively.  Other variables may also be included for each edge, such as the edge weight or grouping variable.  As before, the variables of interest are columns in the data frame and the rows are each edge in the network.

  \subsubsection{Vertex Aesthetics}

@@

\par Evidently, we created all of these aesthetics to fit in exactly to the \texttt{ggplot2} grammar of graphics. We also created the edge aesthetics in a similar fashion.
  \subsubsection{Edge Aesthetics}
  Again, our edge aesthetics mimic the familiar \texttt{ggplot2} aesthetics, this time for segments.    The \texttt{ecolor} aesthetic changes the color of all of the edges according to the identity function or can change the color according to HEX colors assigned to levels of a grouping variable.  In the final version, the \texttt{ecolor} aesthetic will automatically color according to levels of a grouping variable or according to a gradient scale associated with a numeric variable.
  The \texttt{elinetype} aesthetic can also be used with a grouping variable to change the line type of each edge in the graph from solid to one of the different types of lines in \texttt{ggplot2}. Currently, the column associated with this aesthetic must be integer valued and take on values from zero to six. In the final version, it will change the edge linetypes according to levels of a grouping variable.  The \texttt{esize} aesthetic can also be set to the weight or probability of each edge, or it can be changed for all edges with the identity function. The associated column can take on any numerical values, and currently maps the values directly to points, just like the \texttt{vsize} aesthetic. In the final version, this aesthetic will work exactly like the sizing in \texttt{ggplot2}, by scaling sizes to the minimum and maximum values. This is one of two ways to visualize the number of edges between two vertices.  The other way is through the \texttt{ealpha} aesthetic. The \texttt{ealpha} aesthetic can change the edge transparency to a set value on the interval $[0,1]$.  If there are multiple edge connections given between two vertices in the edge data frame, the \texttt{ealpha} aesthetic can be changed to plot opaque lines between vertices with many mutual connections and varying degrees of transparent lines between vertices with only a few or a single connection. This property is also especially useful for large networks with a lot of connections or with vertices of relatively high degree. Finally, we will add an \texttt{elabel} aesthetic to print the name or number assigned to each edge on or next to it.  This aesthetic will be very useful for visualizing random graphs or Markov processes where each edge probability is of interest.

\subsubsection{Other Arguments}
  Similar to the other geometries and functions in \texttt{ggplot2}, our geometry takes several arguments outside of the aesthetic mapping parameters.  One such argument is the \texttt{vlabel} argument. The \texttt{vlabel} aesthetic is a logical value that, if set to \texttt{TRUE} prints the name or number associated with each vertex in the \texttt{label} column in order to identify it on the plot.  In the final version of this geometry, we will add capabilities to adjust the size and positioning of the labels. For the moment, the color is set to the vertex color, the size is set to twice the vertex size, and all other possible parameters are set to their defaults in \texttt{geom\_text()}. The \texttt{vlabel} argument is most useful for smaller network objects where all vertex names can be printed on their corresponding vertices and still be read clearly.  Next, the \texttt{directed} argument is a logical value that identifies whether or not the network is directed.  When this value is true, arrows are created on the end of the line segment that corresponds to the \texttt{to\_id} value for each edge.  The arrows match the edges in coloring, linetype, and size, and the default length of the arrow is set using the \texttt{unit} function in the \texttt{grid} package to 0.015 of the normalised parent coordinates of the plot.
  Finally, the \texttt{layout} argument takes a character value corresponding to the possible layouts in the \texttt{sna} package that are created by the \texttt{gplot.layout.*()} family of functions.  The default layout is the Kamada-Kawai layout.  This is a force-directed layout for undirected networks \citep{kamadakawai}. There are, however, many other layouts possible.
\begin{table}
\centering
\begin{tabular}{l|p{10cm}}
Layout Name & Description \\
\hline \hline
\texttt{"adj"} & a version of \texttt{"mds"} which scales the raw adjacency matrix \\
\texttt{"circle"}* & places vertices uniformly in a circle \\
\texttt{"circrand"} & places vertices randomly in a circle \\
\texttt{"eigen"} & places vertices based on the eigenstructure of the adjacency matrix \\
\texttt{"fruchtermanreingold"} & uses a variant of Fruchterman and Reingold's force-directed placement algorithm \\
\texttt{"geodist"} & a version of \texttt{"mds"} which scales the matrix of geodesic distances \\
\texttt{"hall"}* &  places vertices based on the last two eigenvectors of the Laplacian of the input matrix \\
\texttt{"kamadakawai"} & generates a vertex layout using a version of the Kamada-Kawai force-directed placement algorithm \\
\texttt{"mds"}  & places vertices based on a metric multidimensional scaling of a specified distance matrix \\
\texttt{"random"} & places vertices randomly \\
\texttt{"segeo"} & a version of \texttt{"mds"} which scales the squared euclidean distances between row-wise geodesic distances \\
\texttt{"seham"} & a version of \texttt{"mds"} which scales the Hamming distance between rows/columns of the adjacency matrix \\
\texttt{"target"} & produces a ``target diagram" or ``bullseye" layout using a force-directed placement algorithm.
\end{tabular}
\caption{All possible layouts to pass to \texttt{geom\_net}.  All descriptions are from \citet{sna}.  Note that the layouts marked with an asterisk (*) are the only layouts that do not take layout parameters.}
\label{table:layouts}
\end{table}
